cILR: Taxonomic Enrichment Analysis with Isometric Log-Ratios

### 2021-05-28 tax_agg.tex
Updates from Overleaf

@@

Figure \ref{fig:4} and \ref{fig:4} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks. As expected, performance across all assessed methods increased with a higher signal-to-noise ratio, however, they remained consistent across different levels of model saturation and inter-taxa correlation. Importantly, cILR methods outperformed both GSVA and ssGSEA, where the difference in AUC increases as a function of sparsity. Conversely, cILR scores did not perform as well as the traditional CLR approach, however, the performance gap decreased at higher sparsity levels. Similar patterns were observed for regression results (Figure \ref{fig:4}). Most interestingly, regression performance, in general, decreased at higher saturation levels (proportion of sets associated with the outcome). cILR remained more performant than both ssGSEA and GSVA, but was still less than the standard CLR, especially at higher signal-to-noise ratios. Similarly, cILR still performed much better in higher sparsity situations, where the results were almost identical to the CLR approach, especially in high saturation scenarios.

\begin{figure}[H]
    \centering
    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
    \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio. Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome). cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
    \label{fig:4}
\end{figure}

@@

In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:5}. The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls. The Gevers et al. \cite{gevers2014} data set is a 16S rRNA sequencing data set while the Nielsen et al. \cite{nielsen2014} is a whole-genome shotgun sequencing data set. Similar to simulation experiments, across both data sets cILR methods provide much better performance than either GSVA or ssGSEA. Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set. Additionally, we observed that using the normal distribution for cILR generates better predictive performance compared to the other variants.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
    \caption{Classification performance of a standard random forest model using cILR scores compared against existing methods in gene set testing literature and the standard centered-log ratio transformation approach. The learning task involves predicting patients with inflammatory bowel disease (including Crohn's disease and ulcerative colitis) versus controls. Data sets used span both 16S rRNA sequencing (Gevers et al. \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al. \cite{nielsen2014}). cILR marginally outperforms GSVA and ssGSEA in both data sets, but only performs better than CLR in the WGS data set.}
    \label{fig:5}
\end{figure}

@@

Unsurprisingly, different variants of cILR did not differ significantly in performance similar to significance testing experiments. Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:3}) as well as the random forest evaluations (Figure \ref{fig:4}).

>>>>

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
    \caption{Placeholder} 
    \label{fig:4}
\end{figure}

@@

Figure \ref{fig:6} and \ref{fig:6} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks. As expected, performance across all assessed methods increased with a higher signal-to-noise ratio, however, they remained consistent across different levels of model saturation and inter-taxa correlation. Importantly, cILR methods outperformed both GSVA and ssGSEA, where the difference in AUC increases as a function of sparsity. Conversely, cILR scores did not perform as well as the traditional CLR approach, however, the performance gap decreased at higher sparsity levels. Similar patterns were observed for regression results (Figure \ref{fig:6}). Most interestingly, regression performance, in general, decreased at higher saturation levels (proportion of sets associated with the outcome). cILR remained more performant than both ssGSEA and GSVA, but was still less than the standard CLR, especially at higher signal-to-noise ratios. Similarly, cILR still performed much better in higher sparsity situations, where the results were almost identical to the CLR approach, especially in high saturation scenarios.

\begin{figure}[H]
    \centering
    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
    \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio. Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome). cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
    \label{fig:6}
\end{figure}

@@

In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:7}. The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls. The Gevers et al. \cite{gevers2014} data set is a 16S rRNA sequencing data set while the Nielsen et al. \cite{nielsen2014} is a whole-genome shotgun sequencing data set. Similar to simulation experiments, across both data sets cILR methods provide much better performance than either GSVA or ssGSEA. Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set. Additionally, we observed that using the normal distribution for cILR generates better predictive performance compared to the other variants.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
    \caption{Classification performance of a standard random forest model using cILR scores compared against existing methods in gene set testing literature and the standard centered-log ratio transformation approach. The learning task involves predicting patients with inflammatory bowel disease (including Crohn's disease and ulcerative colitis) versus controls. Data sets used span both 16S rRNA sequencing (Gevers et al. \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al. \cite{nielsen2014}). cILR marginally outperforms GSVA and ssGSEA in both data sets, but only performs better than CLR in the WGS data set.}
    \label{fig:7}
\end{figure}

@@

Unsurprisingly, different variants of cILR did not differ significantly in performance similar to significance testing experiments. Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:6}) as well as the random forest evaluations (Figure \ref{fig:6}).
